Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-21 manu-draft.Rmd
week trend plots to supp, changing file suffix for supp

A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline). After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgamtrnds)b).  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.  A quantitative description of these events follows.

@@

Total nitrogen, chl-a, and secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgamtrnds)).  For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges during April.  Concentrations decreased slightly until June and July when values increased again above the seasonal expectation. Many secchi observations in area 1 were lower than normal in April and July.  Observations in areas 2 and 3 were more often within the normal seasonal range, with some exceptions for total nitrogen and chl-a in area 3 in  April, May, and July.  Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgamtrnds). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1 and for total nitrogen and chl-a for areas 2 and 3 (Table \@ref(tab:wqcmptab)). Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.  Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances.

@@

```{r wqgamtrnds, fig.cap = 'Predicted (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.  Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.  Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.  Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}
knitr::include_graphics(here('figs/wqgamtrnds.jpeg'))
```

>>>>

A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline). After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgam)b).  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.  A quantitative description of these events follows.

@@

Total nitrogen, chl-a, and secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).  For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges during April.  Concentrations decreased slightly until June and July when values increased again above the seasonal expectation. Many secchi observations in area 1 were lower than normal in April and July.  Observations in areas 2 and 3 were more often within the normal seasonal range, with some exceptions for total nitrogen and chl-a in area 3 in  April, May, and July.  Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1 and for total nitrogen and chl-a for areas 2 and 3 (Table \@ref(tab:wqcmptab)). Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.  Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances.

@@

```{r wqgam, fig.cap = 'Predicted (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.  Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.  Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.  Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}
knitr::include_graphics(here('figs/wqgam.jpeg'))
```

### 2021-10-21 supplement.Rmd
week trend plots to supp, changing file suffix for supp

load(file = here('tables/mcrfotab.RData'))
load(file = here('tables/savfotab.RData'))
load(file = here('tables/mcrabutab.RData'))
load(file = here('tables/savabutab.RData'))
```

@@

```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 in response to wastewater discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).  Observations are aggregated by week and within assessment areas shown in Figure 1a.  Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).'}
knitr::include_graphics('figs/wqtrnds-supp.jpeg')
```

```{r wqgamtrnds-supp, fig.cap = 'Predicted (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.  Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.  Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.  Results are grouped by assessment areas shown in Figure 1a.'}
knitr::include_graphics('figs/wqgamtrnds-supp.jpeg')
```

```{r trnabu, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (Figure 1a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.  Estimates are grouped by sample months in 2021. Points are offset slightly for readability.'}
knitr::include_graphics(here('figs/trnabu.jpeg'))
```

```{r windroses, fig.cap = 'Wind rose plots for 2021 by month. Data are from St. Petersburg, Florida.  Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
knitr::include_graphics('figs/windroses.jpeg')
```

```{r phymcrfoest, fig.cap = 'Frequency occurrence estimates for (a) phytoplankton (diatoms and *K. brevis*) and (b) macroalgae groups. Frequency occurrence estimates are aggregated by week of observation based on all sample locations where a phytoplankton or macroalgal taxa was observed divided by all sample locations in a week.  Estimates are not additive and are specific to each taxa.  Sample dates are noted by vertical grey lines in each plot. Diatoms are based on presence/absense of *Asterionellopsis sp*. and *Skeletonema sp*.  Note that frequency occurrence is not a precise measure of cell concentrations.'}
knitr::include_graphics('figs/phymcrfoest.jpeg')
```

@@

```{r mcrfotab}
cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- mcrfotab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savfotab}
cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- savfotab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r mcrabutab}
cap.val <- 'Comparison of macroalgae Braun-Blanquet abundances by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of abundances between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly abundances for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have abundances that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- mcrabutab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savabutab}
cap.val <- 'Comparison of seagrass species Braun-Blanquet abundances by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of abundances between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly abudances for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have abundances that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- savabutab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

>>>>

load(file = here('tables/mcrfotabsupp.RData'))
load(file = here('tables/savfotabsupp.RData'))
load(file = here('tables/mcrabutabsupp.RData'))
load(file = here('tables/savabutabsupp.RData'))
```

@@

```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).  Observations are aggregated by week and within assessment areas shown in Figure 1a.  Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure 1a).'}
knitr::include_graphics(here('figs/wqtrnds-supp.jpeg'))
```

```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 in response to wastewater discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).  Observations are aggregated by week and within assessment areas shown in Figure 1a.  Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).'}
knitr::include_graphics('figs/wqtrndsadd-supp.jpeg')
```

```{r wqgamadd-supp, fig.cap = 'Predicted (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.  Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.  Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.  Results are grouped by assessment areas shown in Figure 1a.'}
knitr::include_graphics('figs/wqgamadd-supp.jpeg')
```

```{r trnabu-supp, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (Figure 1a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.  Estimates are grouped by sample months in 2021. Points are offset slightly for readability.'}
knitr::include_graphics(here('figs/trnabu-supp.jpeg'))
```

```{r windroses-supp, fig.cap = 'Wind rose plots for 2021 by month. Data are from St. Petersburg, Florida.  Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
knitr::include_graphics('figs/windroses-supp.jpeg')
```

```{r phymcrfoest-supp, fig.cap = 'Frequency occurrence estimates for (a) phytoplankton (diatoms and *K. brevis*) and (b) macroalgae groups. Frequency occurrence estimates are aggregated by week of observation based on all sample locations where a phytoplankton or macroalgal taxa was observed divided by all sample locations in a week.  Estimates are not additive and are specific to each taxa.  Sample dates are noted by vertical grey lines in each plot. Diatoms are based on presence/absense of *Asterionellopsis sp*. and *Skeletonema sp*.  Note that frequency occurrence is not a precise measure of cell concentrations.'}
knitr::include_graphics('figs/phymcrfoest-supp.jpeg')
```

@@

```{r mcrfotabsupp}
cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure 1a) and month.  Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- mcrfotabsupp %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savfotabsupp}
cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure 1a) and month.  Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- savfotabsupp %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r mcrabutabsupp}
cap.val <- 'Comparison of macroalgae Braun-Blanquet abundances by areas of interest (Figure 1a) and month.  Overall signifance of differences of abundances between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly abundances for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have abundances that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- mcrabutabsupp %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savabutabsupp}
cap.val <- 'Comparison of seagrass species Braun-Blanquet abundances by areas of interest (Figure 1a) and month.  Overall signifance of differences of abundances between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly abudances for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have abundances that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- savabutabsupp %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

# References
